---
title: "Obama to Trump Voter Switch "
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, message = FALSE)
load("TrumpObama.RData")
library(dplyr)
library(ggplot2)
```

To what extent do attitudes toward immigration explain the switching of votes of 2012 Obama supporters who became 2016 Trump supporters?

We use data from the 2016 Cooperative Congression Election Study to attempt to answer this. The population of interest is taken to be all people who voted for Barack Obama in the 2012 Presidential general election who were eligible to vote in the 2016 Presidential general election. In the first section, we will see to what extent member of this population switching their votes to Donald Trump can be explained by individual demographic variables: gender, education, race, and party identification. We also construct an index to describe attitudes toward immigration and see how this relates to vote switching. In the second section, we test out models with immigration and one other explanatory variable to predict vote switching, looking for interaction. In all cases, we use the CCES survey weights to improve the representativeness of the data.


```{r include = FALSE}
# Question 0
obama <- x[x$CC16_326 == 'Barack Obama' & is.na(x$CC16_326)== FALSE & x$tookpost == 'Yes' ,c('commonweight_vv_post', 'tookpost', 'gender', 'educ', 
                'race', 'pid7', 'CC16_326', 'CC16_410a', 'CC16_331_1',
                                           'CC16_331_2', 'CC16_331_3', 'CC16_331_7')]

obama$votedTrump <- ifelse(obama$CC16_410a == 'Donald Trump (Republican)'& is.na(obama$CC16_410a) == FALSE, 1, 0)

obama$race = recode(obama$race, "White"="White", "Black"="Black", "Hispanic"="Hispanic", .default = "Other" )

obama$CC16_331_1 = recode(obama$CC16_331_1, "Yes"= 1, "No"= 0)
obama$CC16_331_2 = recode(obama$CC16_331_2, "Yes"= 0, "No"= 1)
obama$CC16_331_3 = recode(obama$CC16_331_3, "Yes"= 1, "No"= 0)
obama$CC16_331_7 =recode(obama$CC16_331_7, "Yes"= 0, "No"= 1)

obama$attitude <- obama$CC16_331_1+obama$CC16_331_2+obama$CC16_331_3+obama$CC16_331_7

```

## Univariate summaries

We study the 23,395 people in the 2016 CCES survey who said they voted for Obama for in the 2012 Presidential general election. We find that 10.9% of them (using survey weights) said they voted for Trump in the 2016 Presidential general election. We check how this breaks down by demographic:

**Gender**
```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, message = FALSE)

# For Gender
Total.gender_prop <- obama  %>%  group_by(gender) %>% summarise(SUM = sum(commonweight_vv_post))
gender_prop <- obama  %>% subset(votedTrump == 1) %>%   group_by(gender) %>% summarise(SUM = sum(commonweight_vv_post))

gender_prop$SUM <- gender_prop$SUM/Total.gender_prop$SUM

names(gender_prop) = c("Gender", "Proportion")
``` 

```{r echo = FALSE}
gender_prop
```

**Education**
```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, message = FALSE)

# For Education
Total.educ_prop <- obama  %>%  group_by(educ) %>% summarise(SUM = sum(commonweight_vv_post))
educ_prop <- obama  %>% subset(votedTrump == 1) %>%   group_by(educ) %>% summarise(SUM = sum(commonweight_vv_post))

educ_prop$SUM <- educ_prop$SUM/Total.educ_prop$SUM
```

```{r}
print(educ_prop)
names(educ_prop) = c("Education", "Proportion")
```

**Race**
```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, message = FALSE)

# For Race
Total.race_prop <- obama  %>%  group_by(race) %>% summarise(SUM = sum(commonweight_vv_post))
race_prop <- obama  %>% subset(votedTrump == 1) %>%   group_by(race) %>% summarise(SUM = sum(commonweight_vv_post))

race_prop$SUM <- race_prop$SUM/Total.race_prop$SUM
```

```{r}
print(race_prop)
names(race_prop) = c("Race", "Proportion")
```

**Party Identification**
```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, message = FALSE)
# For pid7
Total.pid7_prop <- obama  %>%  group_by(pid7) %>% summarise(SUM = sum(commonweight_vv_post))
pid7_prop <- obama  %>% subset(votedTrump == 1) %>%   group_by(pid7) %>% summarise(SUM = sum(commonweight_vv_post))

pid7_prop$SUM <- pid7_prop$SUM/Total.pid7_prop$SUM
```

```{r}
print(pid7_prop)
names(pid7_prop) = c("Party Identification", "Proportion")
```

The results line up with conventional wisdom. Male Obama voters were more likely to switch to Trump than female Obama voters, though the gender gap is reasonably small. Obama with a four-year college degree or more education were less likely to switch than those without a four-year college degree. Non-Hispanic white Obama voters were more likely to switch than minorities. Finally, by far the strongest predictor of switching was party identification: those who leaned Republican were likely to vote for Trump even if they voted for Obama, while Democrats were likely to stay loyal. (Of course, a much greater number of Obama voters were Democrats than Republicans.)

We also construct an "immigration attitude" variable from responses to four CCES questions asking what the U.S. government should do about immigration:

- Grant legal status to all illegal immgrants who have held jobs and paid taxes for at least 3 years, and not been convicted of any felony crimes (Yes = 1, No = 0)
- Increase the number of border patrols on the U.S.-Mexican border (No = 1, Yes = 0)
- Grant legal status to people who were brought to the US illegally as children, but who have graduated from a U.S. high school (Yes = 1, No = 0)
- Identify and deport illegal immigrants (No = 1, Yes = 0)

The "immigration attitudes" variable was the sum of these four responses, with 4 being the most positive toward immigration and 0 being the most negative. Note that all these questions are about undocumented immigration, which was the most salient immigration issue during the 2016 campaign; it is possible that asking questions about legal immigration would have substantively different results.

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, message = FALSE)
# For attitude towards immigrangts
Total.attitude_prop <- obama  %>%  group_by(attitude) %>% summarise(SUM = sum(commonweight_vv_post))
attitude_prop <- obama  %>% subset(votedTrump == 1) %>%   group_by(attitude) %>% summarise(SUM = sum(commonweight_vv_post))

attitude_prop$SUM <- attitude_prop$SUM/Total.attitude_prop$SUM
```

```{r}

ggplot(aes(x=attitude,y=SUM), data=attitude_prop) + geom_line() +
  xlab("Attitude Towards Immigrants") + ylab("Weighted Proportion who switched to Trump")

```

## Interactions between immigration attitudes and demographic variables

Before fitting the model, it would be prudent to check the relationship between immigration and the demographic variables.

```{r echo=FALSE,warning=FALSE}
#tookpostsurvey=subset(x,x$tookpost=="Yes")
#voted_Obama=subset(x,x$CC16_326=="Barack Obama")
#respondent_2016=susbset(x,is.na(x$CC16_410a)==FALSE)

#Taking it together

data_preprocess=subset(x,(x$tookpost=="Yes")&(x$CC16_326=="Barack Obama"))
#View(data_preprocess)
respondent_voted_trump=ifelse(data_preprocess$CC16_410a=="Donald Trump (Republican)",1,0)
respondent_voted_trump[which(is.na(respondent_voted_trump))]=0

grant_status_1=ifelse(data_preprocess$CC16_331_1=="Yes",1,0)
grant_status_2=ifelse(data_preprocess$CC16_331_2=="Yes",0,1)
grant_status_3=ifelse(data_preprocess$CC16_331_3=="Yes",1,0)
grant_status_4=ifelse(data_preprocess$CC16_331_7=="Yes",0,1)

attitude=grant_status_1+grant_status_2+grant_status_3+grant_status_4

racial_categories=ifelse(data_preprocess$race=="White","White",ifelse(data_preprocess$race=="Black","Black",ifelse(data_preprocess$race=="Hispanic","Hispanic","Other")))

obama=data.frame(data_preprocess$commonweight_vv_post,data_preprocess$tookpost,data_preprocess$gender,data_preprocess$educ,data_preprocess$race,data_preprocess$pid7,data_preprocess$CC16_326,data_preprocess$CC16_410a,data_preprocess$CC16_331_1,data_preprocess$CC16_331_2,data_preprocess$CC16_331_3,data_preprocess$CC16_331_7,respondent_voted_trump,attitude,racial_categories)



#View(obama)

```



```{r echo=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
#Including interactions and No interactions

#1. With No interactions for Gender demographic variable
gender.logit=glm(respondent_voted_trump~ attitude + data_preprocess.gender,family="binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)

gender.df=expand.grid(attitude=0:4,data_preprocess.gender=c("Male","Female"))

gender.nointer.pred=predict(gender.logit,newdata = gender.df,type = "response")
gender.nointer.pred=as.vector((gender.nointer.pred))

#print(gender.nointer.pred)

#2. With interaction for Gender demographic variable
gender.logit.inter=glm(respondent_voted_trump~attitude*data_preprocess.gender,family = "binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
gender.inter.pred=predict(gender.logit.inter,type = "response",newdata = gender.df)
gender.inter.pred=as.vector((gender.inter.pred))


#Forming a new data frame


gender.pred.df=data.frame(gender.df,gender.nointer.pred,gender.inter.pred)


ggplot(gender.pred.df,aes(x=gender.pred.df$attitude))+geom_line(aes(y=gender.pred.df$gender.nointer.pred,color="No interaction"))+geom_line(aes(y=gender.pred.df$gender.inter.pred,color="with Interaction"))+facet_wrap(~gender.pred.df$data_preprocess.gender)+ylab("Probability to switch to Trump")+xlab("Immigration Attitude")+ggtitle("Interaction of Gender with Immigration attitudes")

```
For the above plots on gender and immigration attitudes interaction we can see that the probabilites to switch to trump predcited by two models(with interaction and without interaction) are same.
Hence gender does not explain any variations in immigration attitude for male and female.


```{r echo=FALSE, warning=FALSE}

#Implementing for race

#1. With No interactions for Race demographic variable
race.logit=glm(respondent_voted_trump~ attitude + data_preprocess.race,family="binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
race.df=expand.grid(attitude=0:4,data_preprocess.race=c("White","Black","Hispanic","Other"))
race.nointer.pred=predict(race.logit,newdata = race.df,type = "response")
race.nointer.pred=as.vector((race.nointer.pred))



#2. With interaction for Race demographic variable
race.logit.inter=glm(respondent_voted_trump~attitude*data_preprocess.race,family = "binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
race.inter.pred=predict(race.logit.inter,type = "response",newdata = race.df)
race.inter.pred=as.vector((race.inter.pred))


#Forming a new data frame


race.pred.df=data.frame(race.df,race.nointer.pred,race.inter.pred)

#Ggplot
ggplot(race.pred.df,aes(x=race.pred.df$attitude))+geom_line(aes(y=race.pred.df$race.nointer.pred,color="No interaction"))+geom_line(aes(y=race.pred.df$race.inter.pred,color="with Interaction"))+facet_wrap(~race.pred.df$data_preprocess.race)+ylab("Probability to switch to Trump")+xlab("Immigration Attitude")+ggtitle("Interaction of Immigration attitudes with Race")


```
From the above plots, we can see that there are interactions of race and immigration attitude variables. Some good differences in probabilities can be seen in Hispanic and Other levels of race. For Hispanic considering the most negative response towards immigration attitude, the probability to switch to trump is slightly less for interaction model as compared to no interaction model.


```{r echo=FALSE, warning=FALSE}
#For Education
#1. With No interactions for Education demographic variable

#Converting quantitative
obama$data_preprocess.educ=as.numeric(obama$data_preprocess.educ)

educ.logit=glm(respondent_voted_trump~ attitude + data_preprocess.educ,family="binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
educ.df=expand.grid(attitude=0:4,data_preprocess.educ=c(1,2,3,4,5,6))
educ.nointer.pred=predict(educ.logit,newdata = educ.df,type = "response")
educ.nointer.pred=as.vector((educ.nointer.pred))



#2. With interaction for Education demographic variable
educ.logit.inter=glm(respondent_voted_trump~attitude*data_preprocess.educ,family = "binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
educ.inter.pred=predict(educ.logit.inter,type = "response",newdata = educ.df)
educ.inter.pred=as.vector((educ.inter.pred))


#Forming a new data frame
educ.pred.df=data.frame(educ.df,educ.nointer.pred,educ.inter.pred)

#Ggplot
ggplot(educ.pred.df,aes(x=educ.pred.df$attitude))+geom_line(aes(y=educ.pred.df$educ.nointer.pred,color="No interaction"))+geom_line(aes(y=educ.pred.df$educ.inter.pred,color="with Interaction"))+facet_wrap(~educ.pred.df$data_preprocess.educ)+ylab("Probability to switch to Trump")+xlab("Immigration Attitude")+ggtitle("Interaction of Immigration attitudes with Education")

```
For 6 levels of education we can see there are some interactions with the immigration attitude.
Some good interactions are shown by the No high school and post grad level of education.
No high School:With interaction we get slight lower probabilites as compared to that of no interaction model for most negative response of immigration attitude.
Post-Grad:With interaction we get slight higher probabilites as compared to that of no interaction model for most negative response of immigration attitude.



```{r echo=FALSE,warning=FALSE}
#For party Identification

#1. With No interactions for Party Identification demographic variable

#Converting quantitative
obama$data_preprocess.pid7=as.numeric(obama$data_preprocess.pid7)

pid.logit=glm(respondent_voted_trump~ attitude + data_preprocess.pid7,family="binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
pid.df=expand.grid(attitude=0:4,data_preprocess.pid7=c(1,2,3,4,5,6,7,8))
pid.nointer.pred=predict(pid.logit,newdata = pid.df,type = "response")
pid.nointer.pred=as.vector((pid.nointer.pred))



#2. With interaction for Education demographic variable
pid.logit.inter=glm(respondent_voted_trump~attitude*data_preprocess.pid7,family = "binomial",data=obama,weights = obama$data_preprocess.commonweight_vv_post)
pid.inter.pred=predict(pid.logit.inter,type = "response",newdata = pid.df)
pid.inter.pred=as.vector((pid.inter.pred))


#Forming a new data frame
pid.pred.df=data.frame(pid.df,pid.nointer.pred,pid.inter.pred)

#Ggplot
ggplot(pid.pred.df,aes(x=pid.pred.df$attitude))+geom_line(aes(y=pid.pred.df$pid.nointer.pred,color="No interaction"))+geom_line(aes(y=pid.pred.df$pid.inter.pred,color="with Interaction"))+facet_wrap(~pid.pred.df$data_preprocess.pid)+ylab("Probability to switch to Trump")+xlab("Immigration Attitude")+ggtitle("Interaction of Immigration attitudes with party Identification")

```
From the above plots we can observe that the party Identification does not make any difference in immigration attitude for predicting the probabilities to switch to trump. There is no interaction between both the variables hence there is no advantage of considering the party Identification demographic variable.

### Comments
In general, the with and without interaction models are very different, indicating that immigration attitudes have a huge amount of predictive power even after accounting for demographic variables. Furthermore, the effect size is massive. For example, race is very well known to be a huge factor in voting, but the models says minorities who were very negative toward immigration were more likely to switch than whites who were very positive toward immigration (keeping gender and party constant.) Immigration attitudes seem to be even more salient for white Obama voters, with male independents who responded negatively to all four CCES immigration questions predicted to be more likely than not to vote for Trump, regardless of education level. We can't prove from this data that negative attitudes toward immigration were *causing* some Obama voters to switch to Trump (since immigration attitudes may be confounded with any number of other variables), but the data is certainly consistent with this story. We also note that positive attitudes toward immigration may have led some voters who would have otherwise switched to refuse to vote for Trump (though even if the effects evened out, it might still be an advantage to Trump due to the peculiarities of the Electoral College.)

## Conclusion
In conclusion, of 2012 Obama voters, Republicans were unsurprisingly the most likely to switch to voting for Trump. Since not many Republicans voted for Obama in the first place and their return to their party may have been inevitable, this may not be a huge deal for the Democratic Party. A bigger concern might be that, according to this data, around a quarter of independent Obama voters switched to Trump. We'd need to compare this to vote switching data from previous elections to see if this was unusual.

Perhaps even more importantly, attitudes toward immigration were a strong predictor of switching to Trump. About 39% of the Obama voters least favorable toward immigration, as measured by CCES responses, switched to Trump, compared to about 2% of the Obama voters most favorable toward immigration. (It would be interesting to do a similar analysis of 2012 Romney voters.) The relationship continues to hold after The data is consistent with immigration attitudes being a stronger determinant of vote switching for white Obama voters, and for Obama voters with at least a college degree. Very speculatively, the Democratic Party might be able to win back such voters without changing policy by making the issue of immigration less salient, though this might result in losing other voters and is easier said than done in any case. For their part, the Republican Party might be well-served by making immigration a major issue for some time to come.



  
